---
title: "A comparison of glider versus satellite-derived SST"
author: Hansen Johnson
date: '2018-10-10'
output: pdf_document
---

```{r setup, include=FALSE}
# This chunk is used to set default options for all the chunks in the entire
# document. In other words, writing 'echo=TRUE' in the function below is the
# same as writing 'echo=TRUE' between the curly brackets at the begining of each
# chunk
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Background

I thought I would take this opportunity to try something I've always wanted to do, which I hope will demonstrate important concepts in BOTH programming *and* oceanography. The **goal** for today will be to compare sea surface temperature (SST) derived from a glider versus a satellite data product from NOAA. This will require a coordinated, thoughtful coding workflow, and (hopefully) shed some light on different types of ocean data.

```{r}

# libraries
library(oce)
library(ncdf4)
library(ocedata)
data("coastlineWorldFine")

# read in functions
source('src/sst_functions.R')
source('src/glider_functions.R')

# define project directory structure
data_dir = 'data/raw/' # raw data
proc_dir = 'data/processed/' # processed data

```

## Datasets

### Glider

Slocum gliders swim up and down through the water column for weeks to months at a time collecting ocean data. The Ocean Tracking Network runs a fleet of these units, which they and various partners use for everything from tracking whales to feeding physical data to assimilatory ocean models.

I'm choosing to plot a live mission that's currently (at the time of this writing) deployed to measure right whale habitat in the southern Gulf of St Lawrence. The `id` for that deployment is `Fundy_20180913_89_realtime`, which is the name of the platform (`Fundy`), the start date (`YYYYmmdd`), OTN mission number (`89`), and type of data (`realtime` for ongoing/live missions, `delayed` for archived/completed missions). You can choose any mission you want on their [website](http://ceotr.ocean.dal.ca/gliders/data). 

The first step is to download and process the data, which I achieve with the custom-written `get_glider_data()` function. You can look at the source code for that function, or any other glider tools, in `src/glider_functions.R`.

```{r}
# define glider deployment id
deployment_id = 'Fundy_20180913_89_realtime'

# get glider data
glider = get_glider_data(deployment_id = deployment_id, data_dir = data_dir, proc_dir = proc_dir)
```

The next step is to visualize the data. I've written functions to plot the CTD data as a simple time series or as a nicely gridded section. The gridded section means that the data are 'normalized' to a grid and smoothed. I'll plot both for comparison. I'll also plot a quick map to get a sense of the glider's spatial coverage.

#### Time series

```{r}
# plot glider time series
plot_time_series(glider)
```

#### Gridded section

```{r}
# plot temperature section
plot_section(glider, var = 'temperature')
```

#### Map

```{r}
# plot coastline
plot(coastlineWorldFine, col = 'grey',
     clon = mean(glider$lon, na.rm = TRUE),
     clat = mean(glider$lat, na.rm = TRUE), 
     span = 200, 
     projection = "+proj=merc")

# add glider lines
mapLines(glider$lon, glider$lat, col = 'blue')

# make text label
txt = paste0(min(glider$time), ' to ', max(glider$time))

# add text
mtext(text = txt, side = 3, line = 0, adj = 0, cex = 0.8)
```

### Satellite

The satellite data I'm using aren't strictly satellite data. It's actually a product from NOAA called the 'Sea surface temperature optimum interpolation' (SSTOI) dataset, which is a compilation of satellite and other remote-sensed data. It's updated daily and has global coverage with 1/4 degree resolution. Most importantly, it's freely available to download! You can learn more [at the NOAA website](). 

The first step is to download the data. I can't resist the opportunity to make a pretty plot, so we might as well plot the full dataset too. Let's begin with an arbitrary date, say the first day of the glider deployment.

```{r}
# test date
td = unique(glider$date)[2]

# read in satellite data
fullsat = get_sst_data(date = td, data_dir = data_dir, proc_dir = proc_dir)

# plot full dataset
plot_sst(fullsat)
```

Beautiful! Now let's extract plot a subset of this data centered around our glider.

```{r}
# find bounding box around glider
bb = calc_bb(glider$lat, glider$lon, lon_buffer = 2, lat_buffer = 1)

# subset satellite
sat = subset_sst(sat = fullsat, bb = bb)

# plot subset sat
plot_sst(sat)
```

### Glider and Satellite

It would be nice to see what the glider track looks like relative to a single grid cell. Let's plot the glider track on the satellite overlay for our day of interest.

```{r}
# plot subset sat
plot_sst(sat, colorbar = FALSE)

# add glider tracks
mapLines(glider$lon, glider$lat, col = 'lightgrey', lwd = 2)

# add tracks for day of interest
with(subset(glider, date == td),
     mapLines(lon, lat, col = 'blue', lwd = 2)
)
```

## Processing 

### Glider 'SST'

Pulling the glider SST is relatively simple. I'll do that by just subsetting the data to only take observations above a given depth threshold (2 meters by default).

```{r}
# set depth threshold
z_min = 2

# extract glider sst
glider_sst = subset(glider, depth < z_min)

# simple time series
plot(glider_sst$date, glider_sst$temperature, xlab = '', ylab = 'Temperature [deg C]')
```

### Satellite SST (near glider)

Perhaps the trickiest step is to extract the satellite SST for the glider's position on a given day. This is achieved by looping through a sequence of days, determining the average glider position, then extracting the satellite SST from the nearest grid cell.

```{r}
# glider start date
d0 = min(glider$date)

# glider end date
d1 = max(glider$date)

# adjust end date for satellite data delay
if(Sys.Date()-d1 < 3){
  message('Omitting glider data from the last 3 days to account for satellite latency')
  
  # redefine end date
  d1 = Sys.Date()-3
  
  # subset glider
  glider_sst = subset(glider_sst, date <= d1)
}

# make sequence of dates
dseq = seq(from = d0, to = d1, by = 1)

# determine satellite temps on each day
SAT = list()
for(ii in seq_along(dseq)){
  
  # define day
  iday = dseq[ii]
  
  # determine glider position
  glat = median(glider$lat[as.character(glider$date) == as.character(iday)], na.rm = TRUE)
  glon = median(glider$lon[as.character(glider$date) == as.character(iday)], na.rm = TRUE)
  
  # download sst data
  isat = get_sst_data(date = iday)
  
  # extract sst at glider
  SAT[[ii]] = extract_sst(mlat = glat, mlon = glon, sat = isat)
  
}

# collapse list to a data frame
sat_sst = do.call(rbind, SAT)

# plot
plot(sat_sst$date, sat_sst$temperature, col = 'blue',  xlab = '', ylab = 'SSTOI [deg C]')
```

## Results

### Time series

Now that we have both datasets, we can compare them any number of ways. I'll start with a simple time series and correlation.

```{r}
# calculate daily means
glider_dmeans = aggregate(temperature ~ date, data = glider_sst, FUN = mean)

# plot all glider data
plot(glider_sst$date, glider_sst$temperature, 
     xlab = '', ylab = 'SST [deg C]', cex = 1, col = 'grey', pch = 1)

# add daily means
points(glider_dmeans$date, glider_dmeans$temperature, 
       col = 'darkslategrey', pch = 16)
lines(glider_dmeans$date, glider_dmeans$temperature, 
       col = 'darkslategrey')

# add satellite
points(sat_sst$date, sat_sst$temperature, 
       col = 'blue', pch = 16)
lines(sat_sst$date, sat_sst$temperature, 
       col = 'blue')

# add legend
legend('topright', cex = 0.8,
       pch = c(1,16,16), 
       lty = c(NA, 1, 1),
       col = c('grey', 'black', 'blue'), 
       legend = c('Glider observations', 'Glider daily means', 'Satellite daily mean')
)
```

### Correlation

```{r}
# build data frame
df = data.frame(glider = glider_dmeans$temperature, sat = sat_sst$temperature)

# plot
plot(df$glider, df$sat, type = 'p', xlab = 'Glider SST [deg C]', ylab = 'Satellite SST [deg C]')
grid()

# correlation
r_cor = cor(x = df$glider, y = df$sat)

# make text string
txt = paste0('Correlation coefficient (r) = ', round(r_cor, 2))

# add to plot
mtext(text = txt, side = 3, line = 0, adj = 0, cex = 0.8)
```

### Bias

```{r}
# calculate bias
sst_diff = sat_sst$temperature - glider_dmeans$temperature

# plot bias
plot(dseq, sst_diff, xlab = '', ylab = 'Difference [sat-glider]')
abline(h = 0, col = 'grey')
```

## Conclusions

There seems to be fantastic agreement between the glider and satellite observations of sea surface temperature (r = `r round(r_cor,2)`), and no evidence of systematic bias in the satellite. The performance of the SSTOI system, at least under these limited circumstances, is impressive. These results frankly surprised me, and, as any good science, have raised more questions than answers. This code could be easily manipulated to conduct this comparison for other glider deployments, or even all deployments. It could also be amended to work with other platforms in other parts of the world. The SSTOI dataset is especially alluring because it includes global coverage all the way back to 1981. The possibilities are endless! Happy coding :)
